---
title: "In-class Exercise 3: Analytical Mapping"

title-block-banner: true

date: "25 january 2023"
date-modified: last-modified

format: html  

execute: 
  message: false
  warning: false

editor: visual
---

## 1. Install and load packages

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
pacman::p_load(sf, tidyverse, tmap)

NGA_wp <- read_rds("data/rds/NGA_wp.rds")
```

## 2. Basic Chloropleth Map

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
p1 <- tm_shape(NGA_wp) + 
  tm_fill("wp_functional",
          n=10, #10 classes
          style="equal", #classification method
          palette="Blues") + #always plural form
  tm_borders(lwd = 0.1, #border thickness
             alpha= 1) +
  tm_layout(main.title = "Distribution of functional water points",
            legend.outside = FALSE)

p2 <- tm_shape(NGA_wp) + 
  tm_fill("total_wp",
          n=10, #10 classes
          style="equal", #classification method
          palette="Blues") + #always plural form
  tm_borders(lwd = 0.1, #border thickness
             alpha= 1) +
  tm_layout(main.title = "Distribution of total water points",
            legend.outside = FALSE)

```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
tmap_arrange(p2,p1,nrow=1)
```

## 3. Chloropleth Maps for Rates

### 3.1 Deriving Proportion of Functional Water Points and Non-Functional Water Points

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
NGA_wp <- NGA_wp %>%
  mutate(pct_functional = wp_functional/total_wp) %>%
  mutate(pct_nonfunctional = wp_nonfunctional/total_wp)

```

### 3.2 Plotting map of rate

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
tm_shape(NGA_wp) + 
  tm_fill("pct_functional",
          n=10, 
          style="equal", 
          palette="Reds") +
  tm_borders(lwd = 0.1, 
             alpha= 1) +
  tm_layout(main.title = "Rate map of functional water points",
            legend.outside = FALSE)

```

## 4. Extreme Value Maps

### 4.1 Percentile Map

### 4.1.1 Data Preparation

Step 1: Exclude records with NA by using the code chunk below.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
#NGA_wp <- NGA_wp %>%
  #drop_na()

NGA_wp <- NGA_wp %>%
  na.omit()
```

### 4.1.2 Why writing functions?

Step 2: Creating customised classification and extracting values

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
percent <- c(0,.01,.1,.5,.9,.99,1)
var <- NGA_wp["pct_functional"] %>%
  st_set_geometry(NULL)
quantile(var[,1], percent)
```

### 4.1.3 Creating the get.var function

Writing a function has three big advantages over using copy-and-paste:

You can give a function an evocative name that makes your code easier to understand. As requirements change, you only need to update code in one place, instead of many. You eliminate the chance of making incidental mistakes when you copy and paste (i.e. updating a variable name in one place, but not in another).

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
get.var <- function(vname,df){
  v <- df[name] %>%
    st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

### 4.1.4 A percentile mapping function

Next, we will write a percentile mapping function by using the code chunk below.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false

percentmap <- function(vnam, df, legtitle=NA, mtitle="Percentile Map"){
  percent <- c(0,.01,.1,.5,.9,.99,1)
  var <- get.var(vnam, df)
  bperc <- quantile(var, percent)
  tm_shape(df) +
  tm_polygons() +
  tm_shape(df) +
     tm_fill(vnam,
             title=legtitle,
             breaks=bperc,
             palette="Blues",
          labels=c("< 1%", "1% - 10%", "10% - 50%", "50% - 90%", "90% - 99%", "> 99%"))  +
  tm_borders() +
  tm_layout(main.title = mtitle, 
            title.position = c("right","bottom"))
}
```

### 4.1.5 Test drive the percentile mapping function

To run the function, type the code chunk as shown below.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
percentmap("total_wp", NGA_wp)

```

### 4.2 Box Map

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
ggplot(data = NGA_wp,
       aes(x = "",
           y = wp_nonfunctional)) +
  geom_boxplot()

```

### 4.2.1 Creating the boxbreaks function

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
boxbreaks <- function(v,mult=1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  # initialize break points vector
  bb <- vector(mode="numeric",length=7)
  # logic for lower and upper fences
  if (lofence < qv[1]) {  # no lower outliers
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) { # no upper outliers
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}

```

### 4.2.2 Creating the get.var function
The code chunk below is an R function to extract a variable as a vector out of an sf data frame.

arguments: vname: variable name (as character, in quotes) df: name of sf data frame returns: v: vector with values (without a column name)

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
get.var <- function(vname,df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}

```

### 4.2.3 Test drive the newly created function
Let’s test the newly created function.
```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
var <- get.var("wp_nonfunctional", NGA_wp) 
boxbreaks(var)
```

### 4.2.4 Boxmap function

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
boxmap <- function(vnam, df, 
                   legtitle=NA,
                   mtitle="Box Map",
                   mult=1.5){
  var <- get.var(vnam,df)
  bb <- boxbreaks(var)
  tm_shape(df) +
    tm_polygons() +
  tm_shape(df) +
     tm_fill(vnam,title=legtitle,
             breaks=bb,
             palette="Blues",
          labels = c("lower outlier", 
                     "< 25%", 
                     "25% - 50%", 
                     "50% - 75%",
                     "> 75%", 
                     "upper outlier"))  +
  tm_borders() +
  tm_layout(main.title = mtitle, 
            title.position = c("left",
                               "top"))
}
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```

### 4.2.5 Recode zero
The code chunk below is used to recode LGAs with zero total water point into NA.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
NGA_wp <- NGA_wp %>%
  mutate(wp_functional = na_if(
    total_wp, total_wp < 0))
```
