---
title: "In-class Ex 7"

title-block-banner: true

date: "6 february 2023" 
date-modified: last-modified 
author: "Rhonda Ho Kah Yee"

format: 
  html: 
    code-fold: true 
    code-tools: true

execute: 
  message: false 
  warning: false 
  eval: false

editor: visual
---

# Getting Started

```{r}
pacman::p_load(sf, sfdep, tmap, tidyverse)
```

# Importing the Data

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan") 

hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
#head(hunan,5) colnames(hunan2012)

hunan_GDPC <- left_join(hunan,hunan2012)%>%
  select(1:4, 7, 15)
```

```{r}
wm_q <- hunan_GDPC %>%
  mutate(nb= st_contiguity(geometry),
         wt = st_weights(nb,
                         style="W"), 
                          #W refers to row standardisation
         .before=1)
```

**Computing global moran'I** (redundant step)

```{r}
moranI <- global_moran(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)
```

**Performing global moran'I test** -\> gives test results and statistics

```{r}
global_moranI <- global_moran_test(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)

global_moranI
```

p-value = 1.095e-06 less that 0.05,

```{r}
global_moranI$I

global_moranI$K
```

We should set a seed number to make our results reproducible.

```{r}

set.seed(1234)
```

**Performing global moran'I test with permutations** The code chunk below intends to run 100 simulations (total simulations = 1+ nsim=99).

```{r}
# with permutations, 

global_moran_perm(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt,
                  nsim=99)
```

It shows 2 sided because it is a large sample test.

**Computing Local Moran'I**

```{r}
lisa <- wm_q %>% 
  mutate(local_moran =local_moran(
    GDPPC, nb, wt, nsim=99),
    .before=1)%>%
  unnest(local_moran) 
  #need to unnest the list to be able to use the values

lisa

#z ii is standardised
# p ii is the permutations

# This step is helpful as we dont hve to manually integrate like we did in the handson exercise (handson ex7 steps 10.7.3 - qunadrant <- vector....)

#general way: mean and pysal(python lib that does the same thing as finding the mean) should be the same

#can use the mean for the takehome assgn , dont hve worry about the median and pysal (feel free to explore)

```

```{r}
tmap_mode("plot")
tm_shape(lisa) +
  tm_fill("p_ii_sim") + #using the simulations is btr than p_ii cos results will be more stable
  tm_borders(alpha=0.5)
  #tm_view(set.zoom.limits=c(6,8))
```

```{r}
# takehome assgn should use gstar instead of lisa 

lisa_sig <-lisa %>%
  filter(p_ii<0.05)

tmap_mode("plot")
tm_shape(lisa) +
  tm_polygons() + 
  tm_borders(alpha=0.5)+

tm_shape(lisa_sig) +
  tm_fill("mean") + 
  tm_borders(alpha=0.4)
  #tm_view(set.zoom.limits=c(6,8))

```

Computing local Moran's I

```{r}
HCSA <- wm_q %>%
  mutate(local_Gi = local_gstar_perm(
    GDPPC, nb, wt, nsim=99),
    .before=1) %>%
  unnest(local_Gi)

HCSA
```

```{r}
tmap_mode("plot")
tm_shape(HCSA) +
  tm_fill("p_sim") +
  tm_borders(alpha=0.05)

# need to make sure to be able to see the significance level
```

```{r}

# GDPPC_nb <- GDPPC_st %>% 
  #activate("geometry") %>%
  #mutate(
   # nb = include_self(st_continuity(geometry)),
    #wt = st_weights(nb)
  #) %>%
  #set_nbs("nb") %>%
  #set_wts("wt")

```

```{r}


```

```{r}

hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")

GDPPC <- read_csv("data/aspatial/Hunan_GDPPC.csv")

```

```{r}
GDPPC_st <- spacetime(GDPPC, hunan,
                      .loc_col = "County",
                      .time_col = "Year")
```

```{r}
is_spacetime_cube(GDPPC_st)
```

```{r}
glimpse(GDPPC)
class(GDPPC$Year)
typeof(GDPPC$Year)
```

```{r}
GDPPC_nb <- GDPPC_st %>%
  activate("geometry") %>%
  mutate(nb = include_self(st_contiguity(geometry)),
         wt = st_inverse_distance(nb, geometry,
                                  scale = 1,
                                  alpha = 1),
         .before = 1) %>%
  set_nbs("nb") %>%
  set_wts("wt")

gi_stars <- GDPPC_nb %>% 
  group_by(Year) %>% 
  mutate(gi_star = local_gstar_perm(
    GDPPC, nb, wt)) %>% 
  tidyr::unnest(gi_star)


cbg <- gi_stars %>% 
  ungroup() %>% 
  filter(County == "Changsha") |> 
  select(County, Year, gi_star)
```
